---
title: "Boston Housing Predictor"
author: "Muki Mdluli"
date: "2026-04-29"
output:
  pdf_document:
    extra_dependencies: ["colortbl", "xcolor", "booktabs", "float", "placeins"]
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "left-align",
  fig.pos = "htbp",
  out.extra = ""
)
```

# Boston Suburbs Housing Price Predictor

This report investigates the key drivers of median home values across Boston 
suburbs using multiple linear regression. The dataset contains 400 suburbs and 
12 predictor variables, ranging from socioeconomic indicators such as the 
proportion of lower-status residents (`lstat`) and crime rate (`crim`), to 
structural characteristics such as average room count (`rm`) and proximity to 
employment centres (`dis`). The analysis follows a structured modelling 
pipeline: exploratory data analysis, full linear regression, backward stepwise 
selection using AIC and BIC criteria, and Ridge and LASSO regularisation. 
Model performance is assessed using 10-fold cross-validation RMSE and 
test-set evaluation on a held-out sample. The goal is to identify the most 
parsimonious and generalisable model that accurately predicts `medv` — the 
median value of owner-occupied homes in \$1,000s.

## Exploration

The first step is to understand the size, structure, missingness, 
distributions, and relationships in the training data. The response variable 
is `medv`, the median value of owner-occupied homes in \$1000s.

A brief description of the predictors is given in Table \ref{tab:vars}.

```{r predictors_defined}
library(kableExtra)

boston_predictors <- data.frame(
  Variable = c("crim", "zn", "indus", "chas", "nox", "rm", "age",
               "dis", "rad", "tax", "ptratio", "lstat", "medv"),
  Description = c(
    "Per capita crime rate by town.",
    "Proportion of residential land zoned for lots over 25,000 sq.ft.",
    "Proportion of non-retail business acres per town.",
    "Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).",
    "Nitrogen oxides concentration (parts per 10 million).",
    "Average number of rooms per dwelling.",
    "Proportion of owner-occupied units built prior to 1940.",
    "Weighted mean of distances to five Boston employment centres.",
    "Index of accessibility to radial highways.",
    "Full-value property-tax rate per $10,000.",
    "Pupil-teacher ratio by town.",
    "Lower status of the population (percent).",
    "Median value of owner-occupied homes in $1000s."
  )
)

boston_predictors |>
  kable(
    caption   = "\\label{tab:vars}Descriptions of predictors in the Boston housing dataset.",
    booktabs  = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped")
  ) |>
  column_spec(2, width = "10cm")
```

A peek into the dataset shows that this training sample has 400 Boston suburbs 
and 13 variables, of which `chas` is a categorical variable representing 
suburbs near the Charles River.

```{r exploration}
boston_data <- read.csv('my_boston.csv', header = TRUE)

boston_data$chas <- factor(
  boston_data$chas,
  levels = c(0, 1),
  labels = c("No", "Yes")
)

proportion_yes <- prop.table(table(boston_data$chas))["Yes"]
complement     <- 1 - proportion_yes
percentage     <- round(complement * 100, 1)
```

The data types of the predictors are shown in Table \ref{tab:types}.

```{r datatype_table}
data_types <- data.frame(
  Variable = names(boston_data),
  Type     = sapply(boston_data, class),
  row.names = NULL
)

data_types |>
  kable(
    caption  = "\\ref{tab:types}Variable types in the Boston housing training data.",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped")
  )
```

The first 10 rows of the dataset are shown in Table \ref{tab:head}.

```{r head_table}
head(boston_data, 10) |>
  kable(
    digits   = 2,
    caption  = "\\label{tab:head}First 10 rows of the Boston housing training data.",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "scale_down")
  )
```

## Visualisations
```{r distributions, fig.cap="\\label{fig:distributions}Histograms with density curves for all variables", fig.width=6, fig.height=4}
numeric_boston      <- boston_data
numeric_boston$chas <- as.numeric(numeric_boston$chas == "Yes")

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))
for (v in names(numeric_boston)) {
  hist(numeric_boston[[v]],
       main        = v,
       xlab        = v,
       col         = "gray85",
       border      = "white",
       probability = TRUE)
  lines(density(numeric_boston[[v]]),
        col = "blue",
        lwd = 2)
}
```
\FloatBarrier
Findings of interest:

- The dataset has no missing values.
- The `chas` variable is very imbalanced: `r percentage`% of suburbs do not 
  bound the Charles River, so predictions involving `chas` are based on 
  relatively few observations.
- Several variables are heavily right-skewed, especially `crim`, `zn`, `rad`, 
  `tax`, `ptratio`, and `lstat`,  as seen in Figure \ref{fig:distributions} suggesting these characteristics are 
  concentrated in certain suburbs rather than evenly distributed across Boston.

### Explanatory-target relationships
```{r medv-relationships, fig.cap="Scatterplots of each explanatory variable against median home value (medv)", fig.width=7, fig.height=6}
scatter_predictors <- c("crim", "zn", "indus", "nox", "rm",
                        "age", "dis", "rad", "tax", "ptratio", "lstat")

par(mfrow = c(3, 4), mar = c(4, 4, 2, 1))
for (v in scatter_predictors) {
  plot(boston_data[[v]], boston_data$medv,
       main = paste("medv vs", v),
       xlab = v,
       ylab = "medv",
       col  = "gray5",
       pch  = 16,
       cex  = 0.55)
  abline(lm(boston_data$medv ~ boston_data[[v]]),
         col = "pink",
         lwd = 2)
}
```

```{r modelling-chas, fig.cap="Boxplots of median home value (medv) by Charles River adjacency (chas)", fig.width=4, fig.height=4}
par(mfrow = c(1, 1))
boxplot(medv ~ chas,
        data = boston_data,
        main = "MEDV by CHAS",
        xlab = "Bounds Charles River",
        ylab = "Median Home Value ($1000s)",
        col  = "lightblue")
```
\FloatBarrier
The clearest single-variable relationships with `medv` are with `lstat` 
(negative), `rm` (positive), and `ptratio` (negative). Median home value 
tends to rise with more rooms and fall as socioeconomic disadvantage and 
pupil-teacher ratios increase.

### Checking for Outliers

```{r outliers, fig.cap="\\label{fig:outliers}Boxplots for each numeric predictor", fig.width=6, fig.height=6}
library(ggplot2)
library(gridExtra)

numeric_boston_no_chas <- numeric_boston[, !names(numeric_boston) %in% "chas"]

boxplots <- lapply(names(numeric_boston_no_chas), function(v) {
  ggplot(numeric_boston_no_chas, aes(y = .data[[v]])) +
    geom_boxplot(fill            = "lightblue",
                 colour          = "black",
                 outlier.colour  = "orange",
                 outlier.size    = 2) +
    labs(title = v, y = v, x = "") +
    theme_minimal()
})

grid.arrange(grobs = boxplots, ncol = 4)
```
\FloatBarrier
The orange points in Figure \ref{fig:outliers} represent outliers, present in `crim`, `zn`, `nox`, `rm`, 
`dis`, `lstat`, and `medv`.

### Multi-variable Correlation
\FloatBarrier
```{r pairs-plot, fig.cap="\\label{fig:pairs}Pairwise scatterplot matrix of the numeric predictors.", fig.width=5, fig.height=5, echo=FALSE}
par(mfrow = c(1, 1))
pairs(
  numeric_boston[, c("crim","zn","indus","chas","nox","rm",
                     "age","dis","rad","tax","ptratio","lstat")],
  col         = "red",
  pch         = 16,
  cex         = 0.45,
  main        = "Pairs Plot: Key Predictors",
  upper.panel = NULL
)
```
\FloatBarrier
The pairwise matrix, (Figure \ref{fig:pairs}) reveals predictors that are strongly related to one 
another. Notable relationships include `age` and `dis`, as well as `rm` and 
`lstat`, suggesting these pairs may act as proxies for each other.

```{r correlations, fig.cap="\\label{fig:corrmatrix}Correlation heatmap", fig.height=6, fig.width=7}
cor_matrix <- cor(numeric_boston)

ord    <- hclust(as.dist(1 - abs(cor_matrix)))$order
cm_ord <- cor_matrix[ord, ord]
n      <- ncol(cm_ord)
pal    <- colorRampPalette(c("navy", "white", "maroon"))(200)

par(mar = c(7, 7, 3, 5))
image(1:n, 1:n, cm_ord[n:1, ],
      axes = FALSE, col = pal, zlim = c(-1, 1),
      xlab = "", ylab = "", main = "Correlation Heatmap")
axis(1, at = 1:n, labels = colnames(cm_ord), las = 2, cex.axis = 0.8)
axis(2, at = 1:n, labels = rev(rownames(cm_ord)), las = 2, cex.axis = 0.8)
text(x      = rep(1:n, each = n),
     y      = rep(1:n, times = n),
     labels = round(cm_ord[n:1, ], 2),
     cex    = 0.65)
box()
legend("right", inset = -0.18,
       legend = c("-1", "0", "1"),
       fill   = c("navy", "white", "maroon"),
       title  = "cor")
```
\FloatBarrier
Figure \ref{fig:corrmatrix} shows the strongest correlations with `medv` are `lstat` ($-0.730$), `rm` ($0.727$), 
`ptratio` ($-0.581$), `indus` ($-0.488$), and `tax` ($-0.484$). Strong 
correlation among predictors — particularly `rad` and `tax` — suggests 
potential multicollinearity in a plain linear model. The exploratory phase 
identifies `lstat`, `rm`, `ptratio`, `tax`, `indus`, `nox`, and `crim` as 
the most promising predictors.

## Model Building

After setting aside approximately 20% of the data for testing, the saturated 
model was fitted to the training data.

```{r modelling-coefficients}
library(broom)
library(kableExtra)

lm_all <- lm(medv ~ ., data = boston_data)

lm_all |>
  tidy() |>
  kable(
    digits   = 4,
    caption  = "\\label{tab:fullreg}Coefficient estimates, standard errors, t-statistics, and p-values for the full linear regression model",
    label    = "fullreg",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "hold_position")
  )
```

### Parameter Significance

From Table \ref{tab:fullreg}, `indus` is the most statistically insignificant 
predictor (p-value $= 0.79$) — holding all other predictors constant, the 
proportion of non-retail business acres does very little to impact median sale 
price. The next least significant predictor is `age` (p-value $= 0.50$), 
indicating that the age of housing stock adds negligible explanatory power 
once other variables are accounted for. Both will be candidates for removal 
during model improvement.

```{r modelling-fit}
lm_summary <- summary(lm_all)

model_stats <- data.frame(
  Statistic = c("Residual standard error",
                "Degrees of freedom",
                "Multiple R-squared",
                "Adjusted R-squared",
                "F-statistic"),
  Value = c(round(lm_summary$sigma, 3),
            lm_summary$df[2],
            round(lm_summary$r.squared, 4),
            round(lm_summary$adj.r.squared, 4),
            round(lm_summary$fstatistic[1], 3))
)

kable(model_stats,
      caption  = "\\label{tab:modelfit}Overall fit statistics for the full linear regression model.",
      booktabs = TRUE) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "hold_position")
  )
```

The residual standard error of `r round(lm_summary$sigma, 3)` indicates that 
on average, the model's predictions deviate from the true median home value 
by approximately \$`r round(lm_summary$sigma * 1000)` on the training data, 
with `r lm_summary$df[2]` degrees of freedom remaining after estimating all 
model parameters. The R\textsuperscript{2} of 
`r round(lm_summary$r.squared, 4)` means that approximately 
`r round(lm_summary$r.squared * 100, 1)`\% of the variation in median home 
values is explained by the model on the training data. The F-statistic of 
`r round(lm_summary$fstatistic[1], 3)` tests the null hypothesis that none of 
the predictors have any linear relationship with `medv`. The extremely small 
associated p-value provides overwhelming evidence to reject this null 
hypothesis, confirming that the model as a whole is statistically significant.

### Testing Predictive Fit

```{r testing, echo=FALSE}
full_boston <- read.csv("boston.csv")

set.seed(1494)
train_ind  <- sample(1:nrow(full_boston), 400, replace = FALSE)
train_data <- full_boston[train_ind, ]
test_data  <- full_boston[-train_ind, ]

test_data$chas <- factor(test_data$chas,
                         levels = c(0, 1),
                         labels = c("No", "Yes"))

y_test    <- test_data$medv
yhat_test <- predict(lm_all, newdata = test_data)

test_mse  <- mean((y_test - yhat_test)^2)
test_rmse <- sqrt(test_mse)
test_r2   <- 1 - sum((y_test - yhat_test)^2) /
                 sum((y_test - mean(y_test))^2)

test_stats <- data.frame(
  Statistic = c("Test MSE", "Test RMSE", "Test R-squared"),
  Value     = c(test_mse, test_rmse, test_r2)
)

kable(test_stats,
      digits    = 4,
      caption   = "\\label{tab:testfit}Test-set performance of the full linear regression model on the held-out Boston suburbs.",
      booktabs  = TRUE,
      col.names = c("Test Statistic", "Value")) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped")
  )
```

The full model explains `r round(test_r2 * 100, 1)`\% of the variation in 
median home values on unseen suburbs (Test R\textsuperscript{2} $=$ 
`r round(test_r2, 4)`), meaning roughly 
`r round((1 - test_r2) * 100, 1)`\% of variation remains unexplained. The 
Test RMSE of `r round(test_rmse, 3)` means that on average, the model's 
predictions are off by approximately \$`r round(test_rmse * 1000)` from the 
true value. The Test MSE of `r round(test_mse, 3)` is the average squared 
prediction error — because errors are squared, larger individual mistakes are 
penalised disproportionately, making the MSE sensitive to the extreme home 
values observed in the right-skewed `medv` distribution. This value serves 
as the baseline benchmark for all subsequent models.

## Model Improvement

To improve the full model, backward stepwise selection was applied using both 
AIC and BIC as stopping criteria. Predictors are removed iteratively starting 
from the least significant, stopping when further removal increases the 
information criterion.

```{r model-improvement}
library(dplyr)
library(broom)
library(kableExtra)

full_model   <- lm(medv ~ ., data = boston_data)
n            <- nrow(boston_data)
backward_AIC <- step(full_model, direction = "backward", trace = 0)
backward_BIC <- step(full_model, direction = "backward", k = log(n), trace = 0)

model_kable <- function(model, cap, lab) {
  model |>
    tidy() |>
    mutate(significant = ifelse(p.value < 0.05, "Yes", "No")) |>
    kable(
      digits    = 4,
      caption   = cap,
      booktabs  = TRUE,
      col.names = c("Predictor", "Estimate", "Std. Error",
                    "t-value", "p-value", "Significant")
    ) |>
    kable_styling(
      full_width    = FALSE,
      latex_options = c("striped")
    )
}
```
```{r full-model-coefficients, echo=FALSE}
model_kable(
  full_model,
  "\\label{tab:fullcoef}Coefficient estimates for the full model.",
  "fullcoef"
)
```

```{r backward-aic-coefficients}
model_kable(
  backward_AIC,
  "\\label{tab:aiccoef}Coefficient estimates for the Backward AIC model",
  "aiccoef"
)
```

```{r backward-bic-coefficients}
model_kable(
  backward_BIC,
  "\\label{tab:biccoef}Coefficient estimates for the Backward BIC model. Both AIC and BIC criteria converged on the same reduced model.",
  "biccoef"
)
```

Both the AIC and BIC backward stepwise procedures removed `indus` and `age`, 
converging on the same 10-predictor model. This agreement between two 
independent criteria strengthens confidence in the variable selection outcome.

### Variable Selection

```{r AIC-BIC-selection}
library(caret)

set.seed(1494)
ctrl <- trainControl(method = "cv", number = 10)

cv_full <- train(formula(full_model),   data = boston_data,
                 method = "lm", trControl = ctrl)
cv_AIC  <- train(formula(backward_AIC), data = boston_data,
                 method = "lm", trControl = ctrl)
cv_BIC  <- train(formula(backward_BIC), data = boston_data,
                 method = "lm", trControl = ctrl)

comparison <- data.frame(
  Model       = c("Full", "Backward AIC", "Backward BIC"),
  Predictors  = c(length(coef(full_model))   - 1,
                  length(coef(backward_AIC)) - 1,
                  length(coef(backward_BIC)) - 1),
  AIC         = round(AIC(full_model, backward_AIC, backward_BIC)$AIC, 2),
  BIC         = round(BIC(full_model, backward_AIC, backward_BIC)$BIC, 2),
  CV_RMSE     = round(c(cv_full$results$RMSE,
                        cv_AIC$results$RMSE,
                        cv_BIC$results$RMSE), 4),
  CV_Rsquared = round(c(cv_full$results$Rsquared,
                        cv_AIC$results$Rsquared,
                        cv_BIC$results$Rsquared), 4)
)

comparison |>
  kable(
    caption  = "Comparison of full and backward stepwise models by number of predictors, AIC, BIC, 10-fold CV RMSE, and CV R-squared. The best model by CV RMSE is highlighted.",
    label    = "modelcomp",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "hold_position")
  ) |>
  row_spec(which.min(comparison$CV_RMSE),
           bold       = TRUE,
           color      = "white",
           background = "green")
```

From Table \ref{tab:modelcomp}, the Backward AIC model achieves the lowest 
CV RMSE of `r round(cv_AIC$results$RMSE, 4)` against the full model's 
`r round(cv_full$results$RMSE, 4)`, making it the preferred model among the 
three. The identical AIC and BIC values for the two backward models confirm 
they selected the same predictors.

### Regularisation

```{r regularization, fig.cap="\\label{fig:regular}10-fold cross-validation MSE curves for Ridge (left) and LASSO (right) regularisation", fig.width=7, fig.height=4}
library(glmnet)
library(ggplot2)
library(gridExtra)

boston_matrix      <- boston_data
boston_matrix$chas <- as.numeric(boston_matrix$chas == "Yes")

X_mat <- model.matrix(medv ~ ., data = boston_matrix)[, -1]
Y_vec <- boston_matrix$medv

set.seed(1494)
cv_ridge  <- cv.glmnet(X_mat, Y_vec, alpha = 0, nfolds = 10)
ridge_mod <- glmnet(X_mat, Y_vec, alpha = 0, lambda = cv_ridge$lambda.min)

cv_lasso  <- cv.glmnet(X_mat, Y_vec, alpha = 1, nfolds = 10)
lasso_mod <- glmnet(X_mat, Y_vec, alpha = 1, lambda = cv_lasso$lambda.min)

ridge_rmse <- sqrt(min(cv_ridge$cvm))
lasso_rmse <- sqrt(min(cv_lasso$cvm))

plot_cv_glmnet <- function(cv_fit, title) {
  df <- data.frame(
    log_lambda = log(cv_fit$lambda),
    mse        = cv_fit$cvm,
    upper      = cv_fit$cvup,
    lower      = cv_fit$cvlo,
    nzero      = cv_fit$nzero
  )
  lambda_min <- log(cv_fit$lambda.min)
  lambda_1se <- log(cv_fit$lambda.1se)
  n_df       <- nrow(df)
  every_n    <- max(1, round(n_df / 15))
  label_df   <- df[seq(1, n_df, by = every_n), ]

  ggplot(df, aes(x = log_lambda, y = mse)) +
    geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15) +
    geom_line(color = "red") +
    geom_point(color = "red", size = 1.5) +
    geom_vline(xintercept = lambda_min, linetype = "dashed") +
    geom_vline(xintercept = lambda_1se, linetype = "dashed") +
    geom_text(data = label_df,
              aes(x = log_lambda, y = max(df$upper), label = nzero),
              size = 3, color = "black") +
    scale_x_continuous(limits = c(0, max(log(cv_fit$lambda)))) +
    labs(title = title,x= expression(log(lambda)), y = "Mean-Squared Error") +
    theme_minimal()
}

p1 <- plot_cv_glmnet(cv_ridge, "Ridge")
p2 <- plot_cv_glmnet(cv_lasso, "LASSO")
grid.arrange(p1, p2, ncol = 2)
```

In Figure \ref{fig:regular} regularisation was applied to control model complexity and guard against 
overfitting. Using `lambda.min` — the penalty value minimising 10-fold CV 
error — Ridge achieved a CV RMSE of `r round(ridge_rmse, 4)` and LASSO 
achieved `r round(lasso_rmse, 4)`. Ridge retains all 12 predictors throughout 
the penalty path, shrinking coefficients toward zero without eliminating them. 
LASSO progressively zeroes out weaker predictors as the penalty increases, 
performing automatic variable selection.

#### Goodness of Model Fit

```{r prediction-and-MSE-calcs}
final_comparison <- data.frame(
  Model   = c("Full", "Backward AIC", "Backward BIC", "Ridge", "LASSO"),
  CV_RMSE = round(c(cv_full$results$RMSE,
                    cv_AIC$results$RMSE,
                    cv_BIC$results$RMSE,
                    ridge_rmse,
                    lasso_rmse), 4)
)

final_comparison |>
  arrange(CV_RMSE) |>
  kable(
    caption  = "Final model comparison by 10-fold cross-validated RMSE across all candidate models. The best-performing model is highlighted.",
    label    = "finalcomp",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "hold_position")
  ) |>
  row_spec(1, bold = TRUE, color = "white", background = "green")
```

From Table \ref{tab:finalcomp}, the Backward AIC model is the best-performing 
model with a CV RMSE of `r round(min(final_comparison$CV_RMSE), 4)`, 
outperforming all alternatives including regularisation methods. Its CV RMSE 
is `r round(final_comparison$CV_RMSE[final_comparison$Model == "Full"] - min(final_comparison$CV_RMSE), 4)` 
lower than the full model, confirming that the removed predictors contributed 
noise rather than signal.

```{r interaction-term}
base_final <- backward_AIC
main_terms <- attr(terms(base_final), "term.labels")

candidate_interactions <- combn(
  main_terms, 2,
  FUN      = function(x) paste(x, collapse = ":"),
  simplify = TRUE
)

set.seed(1494)
fold_id <- sample(rep(1:10, length.out = nrow(boston_data)))

cv_rmse_fn <- function(model_formula) {
  pred <- rep(NA_real_, nrow(boston_data))
  for (k in 1:10) {
    fit_k <- lm(model_formula, data = boston_data[fold_id != k, ])
    pred[fold_id == k] <- predict(fit_k,
                                  newdata = boston_data[fold_id == k, ])
  }
  sqrt(mean((boston_data$medv - pred)^2))
}

score_interaction <- function(interaction_term) {
  f   <- update(formula(base_final), paste(". ~ . +", interaction_term))
  fit <- lm(f, data = boston_data)
  data.frame(
    Interaction   = interaction_term,
    AIC           = AIC(fit),
    BIC           = BIC(fit),
    Adj_R2        = summary(fit)$adj.r.squared,
    CV_RMSE       = cv_rmse_fn(f),
    Test_RMSE     = sqrt(mean((test_data$medv -
                               predict(fit, newdata = test_data))^2)),
    Interaction_p = drop1(fit, test = "F")[interaction_term, "Pr(>F)"]
  )
}

base_row <- data.frame(
  Interaction   = "None",
  AIC           = AIC(base_final),
  BIC           = BIC(base_final),
  Adj_R2        = summary(base_final)$adj.r.squared,
  CV_RMSE       = cv_rmse_fn(formula(base_final)),
  Test_RMSE     = sqrt(mean((test_data$medv -
                             predict(base_final,
                                     newdata = test_data))^2)),
  Interaction_p = NA
)

interaction_results <- do.call(rbind,
                               lapply(candidate_interactions,
                                      score_interaction))
interaction_results <- rbind(base_row, interaction_results)
interaction_results <- interaction_results[
  order(interaction_results$CV_RMSE), ]

best_interaction        <- interaction_results$Interaction[1]
final_interaction_model <- lm(
  update(formula(base_final), paste(". ~ . +", best_interaction)),
  data = boston_data
)
```

The selected interaction term is `r best_interaction`. `rm` measures the 
average number of rooms per dwelling, while `lstat` measures the percentage 
of lower-status residents. The negative `rm:lstat` coefficient indicates that 
the price premium associated with having more rooms is not constant: extra 
rooms add more value in lower-`lstat` suburbs than in higher-`lstat` ones, 
reflecting that spacious homes in affluent areas command a disproportionately 
higher price.

## Residual Diagnostics

```{r residual-diagnostics, fig.cap="Standard residual diagnostic plots for the final interaction model: (top-left) residuals vs fitted values", fig.width=6, fig.height=7}

final_residuals <- residuals(final_interaction_model)
final_fitted    <- fitted(final_interaction_model)
final_std_res   <- rstandard(final_interaction_model)

RSS <- sum(final_residuals^2)
RSE <- sqrt(RSS / df.residual(final_interaction_model))

par(mfrow = c(2, 2))
plot(final_interaction_model, which = 1) # Residuals vs fitted
plot(final_interaction_model, which = 2) # Normal Q-Q
plot(final_interaction_model, which = 3) # Scale-location
plot(final_interaction_model, which = 5) # Residuals vs leverage
par(mfrow = c(1, 1))
```


```{r diagnostic-summary}
diagnostic_summary <- data.frame(
  Diagnostic = c(
    "Mean residual",
    "Residual standard deviation",
    "Largest absolute standardised residual",
    "Observations with |standardised residual| > 2"
  ),
  Value = c(
    round(mean(final_residuals), 3),
    round(sd(final_residuals), 3),
    round(max(abs(final_std_res)), 4),
    sum(abs(final_std_res) > 2)
  )
)

diagnostic_summary |>
  kable(
    caption  = "Numerical residual diagnostic summary for the final interaction model.",
    label    = "diagsum",
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width    = FALSE,
    latex_options = c("striped", "hold_position")
  ) |>
  column_spec(2, width = "3cm")
```

## Conclusion
This analysis set out to predict median home values across Boston suburbs and to 
identify which suburb characteristics most strongly influence housing prices.
Across all modelling approaches evaluated, the Backward AIC model 
with an `r best_interaction` interaction term emerged as the best-performing 
model, achieving the lowest cross-validated RMSE of 
`r round(min(final_comparison$CV_RMSE), 4)`.
The retained predictors in the final model point to three broad categories of 
influence on median home values:

**Socioeconomic status** is the strongest driver. `lstat`, the percentage of 
lower-status residents, has the largest negative relationship with `medv`as
suburbs with higher lower-status residents have lower median home values. 

**Property characteristics** also matter substantially. `rm`, the average 
number of rooms per dwelling, has the strongest positive relationship with 
`medv`. `ptratio`, the pupil-teacher ratio,has a negative effect, suggesting that access to better-resourced schools impacts home prices.

**Environmental and structural factors** such as nitrogen oxide concentration 
(`nox`), highway accessibility (`rad`), property tax rate (`tax`), and crime 
rate (`crim`) all contribute negatively to home values with these houses seeing lower median prices. 
Distance to employment centres (`dis`) has a positive effect, suggesting 
suburbs further from industrial areas are valued more highly.

Notably, `indus` and `age` were dropped by both backward selection criteria, 
confirming that the proportion of non-retail business acres and housing age 
add no meaningful predictive signal once the other variables are accounted for.
The final model performs reasonably well, explaining approximately 
`r round(test_r2 * 100, 1)`\% of variation in median home values on unseen 
suburbs.